Tree establishment in models of forest dynamics: a key priority for further research

### 2022-12-15 manuscript/manuscriptRW.Rmd
corrections in abstract and introduction

Using 15 models and model variants buil to capture longterm (i.e.,
decadal to centennial) forest dynamics at the stand, landscape and
global levels, we simulate tree establishment at 200 sites representing
environmental gradients across Europe. The results are benchmarked
against comprehensive data from forest reserves.

@@

Even though the overestimation of tree establishment represent a
challenge for the usefulness of model~~s~~ projections under uncertain
conditions such a climate change, and although some formulations should
be reconsidered to correct excessive ingrowth levels, it is remarkable
that most of the~~se~~ models capture the essential features of the
highly complex process of tree establishment, while not having been
parameterized with such a data. We conclude that in spite of all the
uncertainties, many models hold promise for assessing longterm forest
dynamics under changing environmental conditions.

@@

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, and (2) assess future tree establishment
e.g., in the context of global change scenarios. Given the current
strategies that are used in models of forest dynamics to represent tree
establishment, model behavior is often facing problems that can have
strong impacts on simulated forest dynamics, such as very high levels of
tree establishment, which then necessitate excess mortality at early
stages of tree life to arrive at correct stand structure and
composition. Also, inadequately high species diversity in tree
establishment may be simulated. This is characteristic of "classical"
forest gap models [@gutiérrez2016], at least as long as the simulations
comprise a multitude of species. Other models use calibration against
local canopy-level data to define appropriate ingrowth levels, which is
likely to strongly hamper the general applicability of these models
e.g., under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species richness due to positive feedback effects such that eventually
just single-species ~~forest~~ stands remain. This is sometimes
corrected by the incorporation of a low level of seed influx of all
species at all times [@lischke2006]. However, simulated species
composition is usually exceedingly sensitive to that level, whereas the
parameters of such functions are poorly constrained by field data.
Lastly, there is often a problem with insufficient observational
constraint on parameter values for models that start from very small
tree sizes (e.g., 10 cm of height or even seeds) and track tree
development in a process-oriented manner by considering a multitude of
ecological influences, rather than emphasizing tree establishment into a
larger size class [e.g., @wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001] if we are to make reliable projections
of future forest dynamics, i.e., when the models need to be operated in
extrapolation mode, but also from a fundamental ecological point of view
(increasing systems understanding). Also in the field of general
ecology, tree establishment has not received a lot of attention
[@hanburybrown_forest_2022], although many sub-processes have been
studied in detail. Yet, little data is available that covers all the
processes within one species along environmental gradients, let alone
for a vast suite of species. Thus, the available data are fragmented,
which constitutes a major problem for model building. On the one hand,
monitoring on permanent sample plots (such as National Forest
Inventories) often measure tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation
phase[@hallsby2015]. Data from permanent plots, although available
across regions, are highly heterogeneous, rendering their utility
difficult in a modeling context[@käber2021]. On the other hand, targeted
studies to measure tree establishment on experimental sites are often
limited in spatial extent, species studied, or the sub-set of processes
that are targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth")
and thus as successful tree establishment. We included models that are
based on a range of "philosophies", from models operating at the stand
to the global scale as well as the range of models from empirically
derived to "process-based" [e.g., @bugmann1996; @fabrika2012;
@mette2014; @lexer2001]. Because of the large variability in tree
establishment patterns in nature and the large number of factors driving
this process --- including some that are not included in most models
such as deer browsing --- we do not aim at a detailed statistical
evaluation of each model. Instead, we aim at evaluating the general
ingrowth patterns and magnitudes simulated by the models, and benchmark
the simulated regeneration niche of multiple species against empirical
data along a wide environmental gradients of temperature, moisture, and
light availability.

>>>>

Using 15 models and model variants built to capture long-term (i.e.,
decadal to centennial) forest dynamics at the stand, landscape and
global levels, we simulate tree establishment at 200 sites representing
environmental gradients across Europe. The results are benchmarked
against comprehensive data from forest reserves.

@@

Even though the overestimation of tree establishment represent a
challenge for the usefulness of model projections under uncertain
conditions such a climate change, and although some formulations should
be reconsidered to correct excessive ingrowth levels, it is remarkable
that most of the models capture the essential features of the highly
complex process of tree establishment, while not having been
parameterized with such a data. We conclude that in spite of all the
uncertainties, many models hold promise for assessing long-term forest
dynamics under changing environmental conditions.

@@

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, and (2) assess future tree establishment
e.g., in the context of global change scenarios. Given the current
strategies that are used in models of forest dynamics to represent tree
establishment, model behavior is often facing problems that can have
strong impacts on simulated forest dynamics, such as very high levels of
tree establishment, which then necessitate excess mortality at early
stages of tree life to arrive at correct stand structure and
composition. Also, inadequately high species diversity in tree
establishment may be simulated. This is characteristic of "classical"
forest gap models [@gutiérrez2016], at least as long as the simulations
comprise a multitude of species. Other models use calibration against
local canopy-level data to define appropriate ingrowth levels, which is
likely to strongly hamper the general applicability of these models
e.g., under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species richness due to positive feedback effects such that eventually
just single-species stands remain. This is sometimes corrected by the
incorporation of a low level of seed influx of all species at all times
[@lischke2006]. However, simulated species composition is usually
exceedingly sensitive to that level, whereas the parameters of such
functions are poorly constrained by field data. Lastly, there is often a
problem with insufficient observational constraint on parameter values
for models that start from very small tree sizes (e.g., 10 cm of height
or even seeds) and track tree development in a process-oriented manner
by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001] if we are to make reliable projections
of future forest dynamics, i.e., when the models need to be operated in
extrapolation mode, but also from a fundamental ecological point of view
(increasing systems understanding). Also in the field of general
ecology, tree establishment has not received a lot of attention
[@hanburybrown_forest_2022], although many sub-processes have been
studied in detail. Yet, little data is available that covers all the
processes within one species along environmental gradients, let alone
for a vast suite of species. Thus, the available data are fragmented,
which constitutes a major problem for model building. On the one hand,
monitoring on permanent sample plots (such as National Forest
Inventories) often measure tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their utility difficult in
a modeling context[@käber2021]. On the other hand, targeted studies to
measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth")
and thus as successful tree establishment. We included models that are
based on a range of "philosophies", from models operating at the stand
to the global scale as well as the range of models from empirically
derived to "process-based" [e.g., @bugmann1996; @fabrika2012;
@mette2014; @lexer2001]. Because of the large variability in tree
establishment patterns in nature and the large number of factors driving
this process --- including some that are not incorporated in most models
such as deer browsing --- we do not aim at a detailed statistical
evaluation of each model. Instead, we aim at evaluating the general
ingrowth patterns and magnitudes simulated by the models, and benchmark
the simulated regeneration niche of multiple species against empirical
data along a wide environmental gradients of temperature, moisture, and
light availability.
